---
title: "Toy examples of fitting mJAM"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pkg-dependencies, include=FALSE}
if(!"susieR" %in% installed.packages()){install.packages("susieR")}
if(!"BayesFactor" %in% installed.packages()){install.packages("BayesFactor")}
if(!"Rmpfr" %in% installed.packages()){install.packages("Rmpfr")}
if(!"tidyverse" %in% installed.packages()){install.packages("tidyverse")}
if(!"matrixcalc" %in% installed.packages()){install.packages("matrixcalc")}
```


## Installation 

Please install the package `hJAM` (Hierarchical Joint Analysis of Marginal Summary Statistics) from github as `mJAM` is developed as part of the `hJAM` package. We're in the process of integrating functions from `hJAM` and those from `mJAM`. 

```{r eval=FALSE}
devtools::install_github("USCbiostats/hJAM")
```


## Special notes for macOS users
For macOS users with apple silicon (M1 Macs), please install R version 4.2-arm64 from https://mac.r-project.org/ . This is a specific build for compatibility with the M1 arm-based Macs. One of crucial dependencies in `mJAM`, `Rmpfr`, is not yet compatible with Intel-based R versions running on M1 Macs and it will cause R/Rstudio to abort. But using version 4.2-arm64 will solve this issue. 

```{r load-pkg}
# source("./Scripts/mJAM_SuSiE_Forward.R")
# source("./Scripts/mJAM_build_CS_v2_med_yty_test.R")
# source("./Scripts/mJAM_logcondp_v3.R")
# source("./Scripts/mJAM_LDpruning.R")
# source("./Scripts/mJAM_Forward.R")
library(hJAM)
library(dplyr) ## for the use of piping operator %>% 
```

## Example 1: Complete SNP information {.tabset}


In the first example, we'll illustrate the usage of `mJAM-SuSiE` and `mJAM-Forward` with a simulated dataset, consisting of 3 ancestry groups and 50 SNPs. The first SNP, named "rs1", is the only true causal SNP.  
 
First, let's load the marginal summary statistics of these 50 SNPs. 

```{r}
# Load marginal betas data 
MargBeta <- read.table("./Example-Data/Marg_1.txt")
names(MargBeta) <- c("SNP", "MargBeta_P1", "MargSEBeta_P1", 
                      "MargBeta_P2", "MargSEBeta_P2",
                      "MargBeta_P3", "MargSEBeta_P3")
head(MargBeta)
```

Then load the effect allele frequencies (not necessarily the minor alleles). 
```{r}
# Load effective allele frequency data
EAF <- read.table("./Example-Data/EAF_1.txt")
names(EAF) <- c("SNP", "EAF_P1", "EAF_P2", "EAF_P3")
head(EAF)
```

Finally one reference dosage matrix for each ancestry group. 
```{r}
# Load individual-level SNP data from reference panel
RefDosage_P1 <- read.table("./Example-Data/Dosage_S1_1.txt", header = TRUE)
RefDosage_P2 <- read.table("./Example-Data/Dosage_S2_1.txt", header = TRUE)
RefDosage_P3 <- read.table("./Example-Data/Dosage_S3_1.txt", header = TRUE)
RefDosage_P1[1:5, 1:5]
RefDosage_P2[1:5, 1:5]
RefDosage_P3[1:5, 1:5]
```

Now we'll use the marginal summary statistics (`beta`, `se`, `EAF` and `reference dosage`) to run `mJAM-SuSiE` and `mJAM-SuSiE`. 


### mJAM-SuSiE

```{r}
# --- Fit mJAM_SuSiE
Ex1_SuSiE_fit <- mJAM_SuSiE(marginal.betas = list(MargBeta$MargBeta_P1, MargBeta$MargBeta_P2, MargBeta$MargBeta_P3), 
                      marginal.se = list(MargBeta$MargSEBeta_P1, MargBeta$MargSEBeta_P2, MargBeta$MargSEBeta_P3), 
                      EAFs = list(EAF$EAF_P1, EAF$EAF_P2, EAF$EAF_P3), 
                      N_GWAS = c(5000, 5000, 5000),
                      X_ref = list(RefDosage_P1,RefDosage_P2,RefDosage_P3),
                      logORToBeta = FALSE,
                      SNP_names = MargBeta$SNP,
                      SuSiE_num_comp = 10, 
                      SuSiE_coverage = 0.95)
## --- mJAM_SuSiE results
mJAM_SuSiE_get_cs(susie_fit = Ex1_SuSiE_fit$fit, SNP_names = MargBeta$SNP)
```

`mJAM_SuSiE` outputs a table with 5 columns: `index`, `coverage`, `CS_size`,`index_SNP` and `CS_SNP`. 

- `index` is a label for a distinct credible set. 
- `coverage` is the empirical coverage of this credible set. 
- `CS_size` shows how many SNPs in total are there in this credible set. 
- `index_SNP` is the name of the index SNP (SNP with highest posterior probability) in this credible set.
- `CS_SNP` is names of individual SNPs selected in this credible set.


### mJAM-Forward
```{r}
# --- Fit mJAM_Forward 
Ex1_Forward_fit <- mJAM_Forward(N_GWAS = c(5000, 5000, 5000), 
                                X_ref = list(RefDosage_P1,RefDosage_P2,RefDosage_P3), 
                                Marg_Result = MargBeta,
                                EAF_Result = EAF, 
                                condp_cut = 0.05/50, # Bonferroni-corrected p-val
                                within_pop_threshold = 0.50, 
                                across_pop_threshold = 0.20,
                                coverage = 0.95,
                                filter_rare = FALSE, 
                                rare_freq = NULL)

```

`mJAM_Forward` outputs a list with three tables: `index`, `cs` and `marginal_est`. 

```{r}
# Here're the selected index SNP(s)
Ex1_Forward_fit$index
```

`index` table lists all the index SNP(s) being selected (`SNP`), along with their log10(p-value) conditional on all SNP(s) above (`cond_log10p`), the log10(p-value) conditional on all other index SNP(s) (`final_log10p`), and the p-value threshold used in this analysis (`pcut`). 

We choose to report log10(p-value) instead of p-value on the original scale because R does not support p-value precision smaller than around 1e-300. To avoid ambiguous results under large-effect regions, we calculate p-values on the log/log10 scales and report these. 

```{r}
# Here're the credible sets SNPs
Ex1_Forward_fit$cs %>% filter(CS_in == T)
```

`cs` table records the posterior probabilities of all SNPs being considered for credible set SNPs. Here we only print out those that are selected in the 95% mJAM credible set (rs1, rs8 and rs10) by filtering this table using `CS_in == TRUE`.  

```{r}
Ex1_Forward_fit$mJAM_marg_est %>% head()
```

`mJAM_marg_est` provides the marginal estimates of all SNPs under the mJAM model. This includes `mJAM_effect` (the point estimate), `mJAM_se` (the standard error of the point estimate) and `mJAM_log10p` (log10 of the p-value based on point estimate and standard error). 


## Example 2: Missing SNP info in some populations {.tabset}
In the second example, we will explore the scenario when we have missing data in marginal summary statistics and in the reference dosage. The dataset used in example 2 is the same as in example 1, except that there are 2 SNPs made missing in either summary statistics or reference dosage: the dosage of rs15 in population 1 is missing; the summary statistic of rs20 in population 3 is missing. 


Now let's load the data: 

```{r}
# Load marginal betas data 
MargBeta <- read.table("./Example-Data/Marg_2.txt", 
                       colClasses = c("character", rep("numeric", 6)))
names(MargBeta) <- c("SNP", "MargBeta_P1", "MargSEBeta_P1", 
                     "MargBeta_P2", "MargSEBeta_P2",
                     "MargBeta_P3", "MargSEBeta_P3")

# Take a look at SNP No.20-25, rs20 in P3 missing
MargBeta[20:25,]
```

```{r}
# Load effective allele frequency data
EAF <- read.table("./Example-Data/EAF_2.txt",
                  colClasses = c("character", rep("numeric", 3)))
names(EAF) <- c("SNP", "EAF_P1", "EAF_P2", "EAF_P3")

# Take a look at SNP No.20-25, rs20 in P3 missing
EAF[20:25, ]
```

Note that rs20's `beta`,`se`, and `EAF` are missing in population 3. 


```{r}
# Load individual-level SNP data from reference panel
RefDosage_P1 <- read.table("./Example-Data/Dosage_S1_2.txt", header = TRUE)
RefDosage_P2 <- read.table("./Example-Data/Dosage_S2_2.txt", header = TRUE)
RefDosage_P3 <- read.table("./Example-Data/Dosage_S3_2.txt", header = TRUE)

# Notice that rs15 in P1 is missing
length(names(RefDosage_P1)); "rs15" %in% names(RefDosage_P1)
length(names(RefDosage_P2)); "rs15" %in% names(RefDosage_P2)
length(names(RefDosage_P3)); "rs15" %in% names(RefDosage_P3)
```

Note that rs15 is missing in the reference panel of population 1. 


**Note on input formatting requirement**:  

When combining summary data from multiple populations, remember to put `NA` into where the SNP information is missing in some populations but present in at least one other population. For SNPs that are missing in ALL populations, just exclude them. 

```{r}
# Fill in rs15 as NAs
RefDosage_P1 <- cbind(RefDosage_P1[,1:14], NA, RefDosage_P1[,15:49])
names(RefDosage_P1) <- names(RefDosage_P2)
```

Now we'll use the marginal summary statistics (`beta`, `se`, `EAF` and `reference dosage`) to run `mJAM-SuSiE` and `mJAM-SuSiE`. 

### mJAM-SuSiE
```{r}
# --- Fit mJAM_SuSiE, with missing SNPs. 
Ex2_SuSiE_fit <- mJAM_SuSiE(marginal.betas = list(MargBeta$MargBeta_P1, MargBeta$MargBeta_P2, MargBeta$MargBeta_P3), 
                       marginal.se = list(MargBeta$MargSEBeta_P1, MargBeta$MargSEBeta_P2, MargBeta$MargSEBeta_P3), 
                       EAFs = list(EAF$EAF_P1, EAF$EAF_P2, EAF$EAF_P3), 
                       N_GWAS = c(5000, 5000, 5000),
                       X_ref = list(RefDosage_P1,RefDosage_P2,RefDosage_P3),
                       logORToBeta = FALSE,
                       p_cases = c(0.5, 0.5, 0.5), 
                       replace_missing = TRUE,
                       SNP_names = MargBeta$SNP,
                       SuSiE_num_comp = 10, 
                       SuSiE_coverage = 0.95)

## --- mJAM_SuSiE results
mJAM_SuSiE_get_cs(susie_fit = Ex2_SuSiE_fit$fit, SNP_names = MargBeta$SNP) 
```

In this toy example, missing values do not affect `mJAM_SuSiE`'s results. 

### mJAM-Forward
```{r}
Ex2_Forward_fit <- mJAM_Forward(N_GWAS = c(5000, 5000, 5000),
                                X_ref = list(RefDosage_P1,RefDosage_P2,RefDosage_P3),
                                Marg_Result = MargBeta,
                                EAF_Result = EAF,
                                condp_cut = 0.05/50, 
                                within_pop_threshold = 0.50,
                                across_pop_threshold = 0.20)


## --- mJAM Forward results: index SNPs  
Ex2_Forward_fit$index

## --- mJAM Forward results: credible set SNPs
Ex2_Forward_fit$cs %>% filter(CS_in == T)

```
In this toy example, missing values do not affect `mJAM_Forward`'s results. 


## session info
```{r}
sessionInfo()
```

