---
title: "Toy examples of fitting mJAM"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r pkg-dependencies, include=FALSE}
if(!"susieR" %in% installed.packages()){install.packages("susieR")}
if(!"BayesFactor" %in% installed.packages()){install.packages("BayesFactor")}
if(!"Rmpfr" %in% installed.packages()){install.packages("Rmpfr")}
if(!"tidyverse" %in% installed.packages()){install.packages("tidyverse")}
if(!"matrixcalc" %in% installed.packages()){install.packages("matrixcalc")}
```


## Installation 

Please install the package `hJAM` (Hierarchical Joint Analysis of Marginal Summary Statistics) from github as `mJAM` is developed as part of the `hJAM` package. The `hJAM` package also contains other `JAM`-family approaches: `hJAM` (for MR and TWAS) and `SHAJAM` (for MR and TWAS with high-dimensional intermediates).  

```{r eval=FALSE}
devtools::install_github("USCbiostats/hJAM")
```


## Special notes for macOS users
For macOS users with apple silicon (M1 Macs), please install R version 4.2-arm64 from https://mac.r-project.org/ . This is a specific build for compatibility with the M1 arm-based Macs. One of crucial dependencies in `mJAM`, `Rmpfr`, is not yet compatible with Intel-based R versions running on M1 Macs and it will cause R/Rstudio to abort. But using version 4.2-arm64 will solve this issue. 

```{r load-pkg}
library(hJAM)
library(dplyr) ## for the use of piping operator %>% 
```

## Example 1: Complete SNP information {.tabset}


In the first example, we'll illustrate the usage of `mJAM-SuSiE` and `mJAM-Forward` with a simulated dataset, consisting of 3 ancestry groups and 50 SNPs. These 50 SNPs consist of 5 LD blocks, each with 10 SNPs and uniform pairwise LD between them. The first SNP, named "rs1", is the only true causal SNP.  
 
First, let's load the marginal summary statistics of these 50 SNPs. 

```{r}
# Load marginal betas data 
MargBeta <- read.table("./Example-Data/Marg_1.txt")
names(MargBeta) <- c("SNP", "MargBeta_P1", "MargSEBeta_P1", 
                      "MargBeta_P2", "MargSEBeta_P2",
                      "MargBeta_P3", "MargSEBeta_P3")
head(MargBeta)
```

Then load the effect allele frequencies (not necessarily the minor alleles). 
```{r}
# Load effective allele frequency data
EAF <- read.table("./Example-Data/EAF_1.txt")
names(EAF) <- c("SNP", "EAF_P1", "EAF_P2", "EAF_P3")
head(EAF)
```

Finally one reference dosage matrix for each ancestry group. 
```{r}
# Load individual-level SNP data from reference panel
RefDosage_P1 <- read.table("./Example-Data/Dosage_S1_1.txt", header = TRUE)
RefDosage_P2 <- read.table("./Example-Data/Dosage_S2_1.txt", header = TRUE)
RefDosage_P3 <- read.table("./Example-Data/Dosage_S3_1.txt", header = TRUE)
RefDosage_P1[1:5, 1:5]
RefDosage_P2[1:5, 1:5]
RefDosage_P3[1:5, 1:5]
```

Now we'll use the marginal summary statistics (`beta`, `se`, `EAF` and `reference dosage`) to run `mJAM-SuSiE` and `mJAM-SuSiE`. 

### mJAM-Forward
```{r}
# --- Fit mJAM_Forward 
Ex1_Forward_fit <- mJAM_Forward(N_GWAS = c(5000, 5000, 5000), 
                                X_ref = list(RefDosage_P1,RefDosage_P2,RefDosage_P3), 
                                Marg_Result = MargBeta,
                                EAF_Result = EAF, 
                                condp_cut = 0.05/50, # Bonferroni-corrected p-val
                                within_pop_threshold = 0.50, 
                                across_pop_threshold = 0.20,
                                coverage = 0.95)

```

`mJAM_Forward` outputs a list with three tables: `index`, `cs` and `marginal_est`. 

```{r}
# Here're the selected index SNP(s)
Ex1_Forward_fit$index
```

`index` table lists all the index SNP(s) being selected (`SNP`), along with their log10(p-value) conditional on all SNP(s) above (`cond_log10p`), the log10(p-value) conditional on all other index SNP(s) (`final_log10p`), and the p-value threshold used in this analysis (`pcut`). `b_joint` is the conditional effect estimates in the final model, and `b_joint_var` is the squared standard error of `b_joint`.  

We choose to report log10(p-value) instead of p-value on the original scale because R does not support p-value precision smaller than around 1e-300. To avoid ambiguous results under large-effect regions, we calculate p-values on the log/log10 scales and report these. 

```{r}
# Here're the credible sets SNPs
Ex1_Forward_fit$cs %>% filter(CS_in == T)
```

`cs` table records the posterior probabilities of all SNPs being considered for credible set SNPs. `Post_Model_Prob` stands for posterior model probability which quantifies the marginal association between `CS_SNP` and its `index_SNP`. `Post_Med_Prob` stands for posterior mediation probability which quantifies the mediation effect of the `CS_SNP` through `index_SNP`. Stronger LD between `CS_SNP` and `index_SNP` often leads to larger `Post_Med_Prob`. By taking the product of `Post_Model_Prob` and `Post_Med_Prob` and then standardizing, we have `SD_Post_CS_Prob` which we use to determine the credible set.  

Here we only print out those that are selected in the 95% mJAM credible set (rs1, rs8 and rs10) by filtering this table using `CS_in == TRUE`.  

```{r}
Ex1_Forward_fit$mJAM_marg_est %>% head()
```

If you'd like to check out the multi-population marginal estimates from mJAM, `mJAM_marg_est` provides the marginal estimates of all SNPs under the mJAM model. This includes `mJAM_effect` (the point estimate), `mJAM_se` (the standard error of the point estimate) and `mJAM_log10p` (log10 of the p-value based on point estimate and standard error). 


### mJAM-SuSiE

```{r}
# --- Fit mJAM_SuSiE
Ex1_SuSiE_fit <- mJAM_SuSiE(marginal.betas = list(MargBeta$MargBeta_P1, MargBeta$MargBeta_P2, MargBeta$MargBeta_P3), 
                      marginal.se = list(MargBeta$MargSEBeta_P1, MargBeta$MargSEBeta_P2, MargBeta$MargSEBeta_P3), 
                      EAFs = list(EAF$EAF_P1, EAF$EAF_P2, EAF$EAF_P3), 
                      N_GWAS = c(5000, 5000, 5000),
                      X_ref = list(RefDosage_P1,RefDosage_P2,RefDosage_P3),
                      SNP_names = MargBeta$SNP,
                      SuSiE_num_comp = 10, 
                      SuSiE_coverage = 0.95)
## --- mJAM_SuSiE results
mJAM_SuSiE_get_cs(susie_fit = Ex1_SuSiE_fit$fit, SNP_names = MargBeta$SNP)
```

`mJAM_SuSiE` outputs a table with 5 columns: `index`, `coverage`, `CS_size`,`index_SNP` and `CS_SNP`. 

- `index` is a label for a distinct credible set. 
- `coverage` is the empirical coverage of this credible set. 
- `CS_size` shows how many SNPs in total are there in this credible set. 
- `index_SNP` is the name of the index SNP (SNP with highest posterior probability) in this credible set.
- `CS_SNP` is names of individual SNPs selected in this credible set.

In this example, mJAM-SuSiE identifies one credible set with `rs1` and `rs8` in it. Both of them are included in the mJAM-Forward credible set as well. 


## Example 2: Missing SNP info in some populations {.tabset}
In the second example, we will explore the scenario when we have missing data in marginal summary statistics and in the reference dosage. The dataset used in example 2 is the same as in example 1, except that there are 2 SNPs made missing in either summary statistics or reference dosage: the dosage of rs15 in population 1 is missing; the summary statistic of rs20 in population 3 is missing. 


Now let's load the data: 

```{r}
# Load marginal betas data 
MargBeta <- read.table("./Example-Data/Marg_2.txt", 
                       colClasses = c("character", rep("numeric", 6)))
names(MargBeta) <- c("SNP", "MargBeta_P1", "MargSEBeta_P1", 
                     "MargBeta_P2", "MargSEBeta_P2",
                     "MargBeta_P3", "MargSEBeta_P3")

# Take a look at SNP No.20-25, rs20 in P3 missing
MargBeta[20:25,]
```

```{r}
# Load effective allele frequency data
EAF <- read.table("./Example-Data/EAF_2.txt",
                  colClasses = c("character", rep("numeric", 3)))
names(EAF) <- c("SNP", "EAF_P1", "EAF_P2", "EAF_P3")

# Take a look at SNP No.20-25, rs20 in P3 missing
EAF[20:25, ]
```

Note that rs20's `beta`,`se`, and `EAF` are missing in population 3. 


```{r}
# Load individual-level SNP data from reference panel
RefDosage_P1 <- read.table("./Example-Data/Dosage_S1_2.txt", header = TRUE)
RefDosage_P2 <- read.table("./Example-Data/Dosage_S2_2.txt", header = TRUE)
RefDosage_P3 <- read.table("./Example-Data/Dosage_S3_2.txt", header = TRUE)

# Notice that rs15 in P1 is missing
length(names(RefDosage_P1)); "rs15" %in% names(RefDosage_P1)
length(names(RefDosage_P2)); "rs15" %in% names(RefDosage_P2)
length(names(RefDosage_P3)); "rs15" %in% names(RefDosage_P3)
```

Note that rs15 is missing in the reference panel of population 1. 


**Note on input formatting requirement**:  

When combining summary data from multiple populations, remember to put `NA` into where the SNP information is missing in some populations but present in at least one other population. For SNPs that are missing in ALL populations, just exclude them. 

```{r}
# Fill in rs15 as NAs
RefDosage_P1 <- cbind(RefDosage_P1[,1:14], NA, RefDosage_P1[,15:49])
names(RefDosage_P1) <- names(RefDosage_P2)
```

Now we'll use the marginal summary statistics (`beta`, `se`, `EAF` and `reference dosage`) to run `mJAM-SuSiE` and `mJAM-SuSiE`. 


### mJAM-Forward
```{r}
Ex2_Forward_fit <- mJAM_Forward(N_GWAS = c(5000, 5000, 5000),
                                X_ref = list(RefDosage_P1,RefDosage_P2,RefDosage_P3),
                                Marg_Result = MargBeta,
                                EAF_Result = EAF,
                                condp_cut = 0.05/50, 
                                within_pop_threshold = 0.50,
                                across_pop_threshold = 0.20)


## --- mJAM Forward results: index SNPs  
Ex2_Forward_fit$index

## --- mJAM Forward results: credible set SNPs
Ex2_Forward_fit$cs %>% filter(CS_in == T)

```
In this toy example, missing values do not affect `mJAM_Forward`'s results. 

### mJAM-SuSiE
```{r}
# --- Fit mJAM_SuSiE, with missing SNPs. 
Ex2_SuSiE_fit <- mJAM_SuSiE(marginal.betas = list(MargBeta$MargBeta_P1, MargBeta$MargBeta_P2, MargBeta$MargBeta_P3), 
                       marginal.se = list(MargBeta$MargSEBeta_P1, MargBeta$MargSEBeta_P2, MargBeta$MargSEBeta_P3), 
                       EAFs = list(EAF$EAF_P1, EAF$EAF_P2, EAF$EAF_P3), 
                       N_GWAS = c(5000, 5000, 5000),
                       X_ref = list(RefDosage_P1,RefDosage_P2,RefDosage_P3),
                       p_cases = c(0.5, 0.5, 0.5), 
                       SNP_names = MargBeta$SNP,
                       SuSiE_num_comp = 10, 
                       SuSiE_coverage = 0.95)

## --- mJAM_SuSiE results
mJAM_SuSiE_get_cs(susie_fit = Ex2_SuSiE_fit$fit, SNP_names = MargBeta$SNP) 
```

In this toy example, missing values do not affect `mJAM_SuSiE`'s results. 


## Example 3: mJAM-Forward with pre-specified index

One practical advantage of mJAM-Forward is that it provides the flexibility to pre-specify your own index SNP by using the option `index_snps` in `mJAM_Forward()`. This will bypass the step of index SNP selection but still construct credible sets of the specified index SNPs. In the next example, we will specify two imaginary index SNPs: `rs2` and `rs11` by adding `index_snps = c("rs2", "rs11")` to `mJAM_Forward()`.  

Note that the order of `index_snps` matters: if we put `index_snps = c("rs2", "rs11")`, `mJAM_Forward` will first construct the credible set for `rs2` and then prune out SNPs highly correlated with `rs2`. Then it will construct the credible set for `rs11` and form credible set of `rs11` by conditioning on the effect of `rs2`. If you'd like to get unconditional credible sets for `rs2` and `rs11` separately, we recommend running `mJAM_Forward` twice where specifying one `index_snp` at a time. 

```{r}
Ex3_Forward_fit <- mJAM_Forward(N_GWAS = c(5000, 5000, 5000),
                                X_ref = list(RefDosage_P1,RefDosage_P2,RefDosage_P3),
                                Marg_Result = MargBeta,
                                EAF_Result = EAF,
                                condp_cut = 0.05/50, 
                                index_snps = c("rs2", "rs11"), 
                                within_pop_threshold = 0.50,
                                across_pop_threshold = 0.20)


## --- mJAM Forward results: index SNPs  
Ex3_Forward_fit$index

## --- mJAM Forward results: credible set SNPs
Ex3_Forward_fit$cs %>% filter(CS_in == T) 
```
In the above example, mJAM credible set for `rs2` consists of all 10 SNPs in the same LD block as `rs2`. Conditional on `rs2`, the credible set for `rs11` consists of 5 SNPs (all within the same LD block as `rs11`). 

## session info
```{r}
sessionInfo()
```

